A geometric relationship of $F_2, F_3$ and $F_4$-statistics with Principal Component Analysis

### 2022-03-07 fstats_pca_lewontin.tex
final changes

The two data sets I analyzed here suggest that two PCs (for the World data set) respectively three PCs (for the Western Eurasian data) already provide a very good approximation for $F_4$-statistics (Figures \ref{fig:weu}E, \ref{fig:world}E), reflecting the observation that frequently the first few PCs provide a good approximation of population structure. Thus, we may expect that whenever the first few PCs provide a good approximation of population structure, this results in well-approximated $F_4$-statistics. On the other hand, for admixture $F_3$-statistics, more PCs are needed (Figure \ref{fig:weu}B), likely because these statistics do contain a term measuring local variation in the putative admixed population \citep{peter2016}, which is typically often incorporated on higher PCs. Thus, I would also expect other $F$-statistics including large contribution from local variation such as  $F_4$-statistics between closely related populations to require substantially more PCs.

@@

\subsubsection{Missing data}
The matrix decompositions underlying PCA assume complete data, and thus cannot be used when some data is missing \citep{jolliffe2013}.  As missing data is a very common practical problem, there is a large number of algorithms for imputing missing data. The simplest approach is to replace missing data with zeros \citep[as implemented e.g. in][]{patterson2006}, but more sophisticated algorithms exist to ``learn'' the missing values from surrounding data \citep[e.g.][]{hastie2015, meisner2018, meisner2021}. In contrast, missing data in $F$-statistics is most commonly handled by estimating a standard error by resampling along the genome \citep{patterson2012}, and so missing data results in larger standard errors.

@@

Thus, a more abstract way of interpreting $F$-statistics is that they reflect the relative position of population to each other in PCA-space. The classical model of admixture as a result of contact between long-separated and isolated populations is just one of potentially many demographic models that might result in this data pattern.

\begin{comment}
This continuum between admixture and population structure becomes more problematic when interpreting tools that  use multiple $F$-statistics to build compound models, such as \texttt{qpGraph} \citep{lazaridis2014} and \texttt{qpAdm} \citep{harney2021}. One issue with these approaches is that they are usually restricted to at most a few dozen populations. As ancient DNA data sets now commonly include thousands of individuals, analysts are faced with the challenge of which data to include. A common approach is to sample a large number of distinct models, and retain the ones that are compatible with the data. However, as both \texttt{qpGraph} and \texttt{qpAdm} assume that gene flow is rare and discrete, selecting sets of populations that did experience little gene flow will provide good fits. The PCA-based  interpretation of $F$-statistics offers an alternative that trades interpretability for robustness, as I do not need to assume that gene flow is rare.
\end{comment}

In this vein, admixture tests based on $F$-statistics evaluate if drift in different parts of the tree are orthogonal to each other. This orthogonality is expected to arise whenever drift in different populations is independent, which happens, for example, if we consider two isolated populations, or if we consider the changes in allele frequency in distinct time intervals.

In the context of tree-like population models, this orthogonality arises from drift on non-overlapping branches.

an alternative could be to focus on orthogonality of drift as a \emph{the} central concept for analyzing genetic variation data.

This is often informally done in the context of PCA, where it is often observed that a particular PC is associated with variation within a specific region, or separates out distinct populations \citep{cavalli-sforza1994}. These associations are not exact, and impacted e.g. by the size and composition of the analyzed data set. On the other hand, we can use $F$-statistics to formally test orthogonality. Distinct clades in an admixture graph would be one possibility how orthogonality could arise, but assuming a graph is not necessary. Motivated by PCA, we could set up $F_4$ as tests of orthogonality due to either space (distinct populations evolve independently) or scale (i.e. between-population diversity is independent of within-population diversity). This slight generalization of $F$-statistics could allow us to reframe many of the questions that are currently asked as tests of orthogonality without assuming that lineages or admixture events are discrete.

>>>>

The two data sets I analyzed here suggest that two PCs (for the World data set) respectively three PCs (for the Western Eurasian data) already provide a very good approximation for $F_4$-statistics (Figures \ref{fig:weu}E, \ref{fig:world}E), reflecting the observation that frequently the first few PCs provide a good approximation of the overall population structure. On the other hand, for admixture $F_3$-statistics, more PCs are needed (Figure \ref{fig:weu}B). This is likely due to PCA approximating the global structure in the data set; statistics that only involve distantly related populations will only require a few PCs for good approximations, whereas statistics that contain a term measuring local variation, such as $F_3$-statistics or $F_4$ between closely related populations will require more PCs for good approximations, because local variation is often found on higher PCs.

@@

\subsubsection{Missing data}
The matrix decompositions underlying PCA assume complete data, and thus cannot be used when some data is missing \citep{jolliffe2013}.  As missing data is a very common practical problem, there is a large number of algorithms for imputing missing data. The simplest approach is to replace missing data with zeros \citep[as implemented e.g. in][]{patterson2006}, but more sophisticated algorithms exist to ``learn'' the missing values from surrounding data \citep[e.g.][]{hastie2015,  meisner2021}. In contrast, missing data in $F$-statistics is most commonly handled by estimating a standard error by resampling along the genome \citep{patterson2012}, and so missing data results in larger standard errors.

@@

Thus, a more general way we could think about modeling population structure using $F$-statistics is as identifying orthogonal drift components. In a tree model, orthogonality arises  because changes in allele frequencies on distinct branches of the tree are independent from each other (and high-dimensional random vectors are almost surely orthogonal).  The classical model of admixture as a result of contact between long-separated and isolated populations is one of potentially many demographic models that results in non-orthogonality; gene flow of any kind will result in correlated genetic drift, and hence in non-zero $F_3$ or $F_4$-statistics.

Thinking of population structure in terms of orthogonal components may be abstract, but it is quite similar to how PCA is sometimes interpreted. In a PCA, we frequently make the informal observation that a particular PC is associated with variation within a specific region, or separates out distinct populations \citep{cavalli-sforza1994}. These associations are not exact, and impacted e.g. by the size and composition of the analyzed data set. On the other hand, we could use $F$-statistics to formally test orthogonality, or to quantify correlations.  Motivated by PCA, we could set up $F_4$ as tests of orthogonality due to either space (distinct populations evolve independently) or scale (i.e. between-population diversity is independent of within-population diversity). This slight generalization of $F$-statistics could allow us to reframe many questions about gene flow or divergence that are currently asked as tests of orthogonality, without assuming that lineages or admixture events are discrete.
